---
output:
  pdf_document: default
  html_document: default
---
<center>

<h1>

STAT 360: Computational Statistics and Data Analysis

<h2>Project - DAT</h2>

<h4>Matt Lao & Adam Bihi</h4>

</center>

### Load Relevant R Libraries

```{r warning=FALSE, message=FALSE}
library(rmarkdown); library(knitr); library(moments); 
library(scatterplot3d); library(corrplot); library(pso)
library(psych); library(GPArotation); library(lavaan); library(readxl)
```

### Matrix Construction

```{r}
data_matrix_csv <- read.csv("C:/Users/User/OneDrive - University of St. Thomas/Classes/STAT360/STAT360 Project/data/Data Matrix.csv", row.names=1)
life_expectancy <- as.matrix(data_matrix_csv)

col_names_unedited <- c(
  "Life expectancy at birth, total (years)",
  "CO2 emissions (metric tons per capita)",
  "Access to electricity (% of population)",
  "Current health expenditure (% of GDP)",
  "Out-of-pocket expenditure (% of current health expenditure)",
  "Domestic private health expenditure per capita, PPP (current international $)",
  "Domestic general government health expenditure per capita, PPP (current international $)",
  "Renewable internal freshwater resources per capita (cubic meters)",
  "Prevalence of HIV, total (% of population ages 15-49)",
  "Unemployment, total (% of total labor force) (national estimate)",
  "Government Effectiveness: Estimate",
  "Income share held by highest 10%",
  "Prevalence of current tobacco use (% of adults)",
  "Total alcohol consumption per capita (liters of pure alcohol, projected estimates, 15+ years of age)",
  "Political Stability and Absence of Violence/Terrorism: Estimate",
  "Population density (people per sq. km of land area)"
)

# Better for displaying
col_names <- c(
  "Life_expectancy",
  "CO2_emissions",
  "Electricity",
  "Health_expenditure",
  "Out_of_pocket",
  "Private_health_expenditure",
  "Govt_health_expenditure",
  "Freshwater_resources",
  "HIV_prevalence",
  "Unemployment",
  "Govt_effectiveness",
  "Income_share",
  "Tobacco_use",
  "Alcohol_consumption",
  "Political_stability",
  "Population_density"
)

colnames(life_expectancy) <- col_names
```

### Matrix Output

```{r}
head(life_expectancy)
dim(life_expectancy)
```

**Response:** "Life expectancy at birth, total (years)"

**Predictors:**

"CO2 emissions (metric tons per capita)",

"Access to electricity (% of population)",

"Current health expenditure (% of GDP)",

"Out-of-pocket expenditure (% of current health expenditure)",

"Domestic private health expenditure per capita, PPP (current international \$)",

"Domestic general government health expenditure per capita, PPP (current international \$)",

"Renewable internal freshwater resources per capita (cubic meters)",

"Prevalence of HIV, total (% of population ages 15-49)",

"Unemployment, total (% of total labor force) (national estimate)",

"Government Effectiveness: Estimate",

"Income share held by highest 10%",

"Prevalence of current tobacco use (% of adults)",

"Total alcohol consumption per capita (liters of pure alcohol, projected estimates, 15+ years of age)",

"Political Stability and Absence of Violence/Terrorism: Estimate",

"Population density (people per sq. km of land area)"

### Adjustments to Original Dimensions

-   No adjustments needed at this time
