---
output:
  pdf_document: default
  html_document: default
---
<center>

<h1>

STAT 360: Computational Statistics and Data Analysis

<h2>Project - EFA</h2>

<h4>Matt Lao & Adam Bihi</h4>

</center>

### Load Relevant R Libraries

```{r warning=FALSE, message=FALSE}
library(rmarkdown); library(knitr); library(moments); 
library(scatterplot3d); library(corrplot); library(pso)
library(psych); library(GPArotation); library(lavaan); library(readxl); library(dplyr)
```

### Matrix Construction

```{r}
data_matrix_csv <- read.csv("C:/Users/User/OneDrive - University of St. Thomas/Classes/STAT360/STAT360 Project/data/Data Matrix.csv", row.names=1)
life_expectancy <- as.matrix(data_matrix_csv)

col_names_unedited <- c(
  "Life expectancy at birth, total (years)",
  "CO2 emissions (metric tons per capita)",
  "Access to electricity (% of population)",
  "Current health expenditure (% of GDP)",
  "Out-of-pocket expenditure (% of current health expenditure)",
  "Domestic private health expenditure per capita, PPP (current international $)",
  "Domestic general government health expenditure per capita, PPP (current international $)",
  "Renewable internal freshwater resources per capita (cubic meters)",
  "Prevalence of HIV, total (% of population ages 15-49)",
  "Unemployment, total (% of total labor force) (national estimate)",
  "Government Effectiveness: Estimate",
  "Income share held by highest 10%",
  "Prevalence of current tobacco use (% of adults)",
  "Total alcohol consumption per capita (liters of pure alcohol, projected estimates, 15+ years of age)",
  "Political Stability and Absence of Violence/Terrorism: Estimate",
  "Population density (people per sq. km of land area)"
)

# Better for displaying
col_names <- c(
  "Life_expectancy",
  "CO2_emissions",
  "Electricity",
  "Health_expenditure",
  "Out_of_pocket",
  "Private_health_expenditure",
  "Govt_health_expenditure",
  "Freshwater_resources",
  "HIV_prevalence",
  "Unemployment",
  "Govt_effectiveness",
  "Income_share",
  "Tobacco_use",
  "Alcohol_consumption",
  "Political_stability",
  "Population_density"
)

colnames(life_expectancy) <- col_names



```

### Variance-Covariance Matrix

```{r}
SIGMA <- cov(life_expectancy, use = "pairwise.complete.obs")
SIGMA
```

### Covariance Matrix (Response/Predictors)

```{r}
SIGMA_p <- SIGMA[2:16,2:16]
SIGMA_p

SIGMA_r  <- SIGMA[1,1]
SIGMA_r  <- as.matrix(SIGMA_r)
SIGMA_r
```

### Correlation Matrix (Predictor/Response)

```{r}
R_p <- (solve(sqrt(diag(diag(SIGMA_p))))) %*% SIGMA_p %*%
t(solve(sqrt(diag(diag(SIGMA_p)))))
dimnames(R_p) <- list(c("CO2_emissions", "Electricity", "Health_expenditure", "Out_of_pocket", "Private_health_expenditure",
  "Govt_health_expenditure","Freshwater_resources","HIV_prevalence","Unemployment", "Govt_effectiveness",
  "Income_share","Tobacco_use","Alcohol_consumption", "Political_stability", "Population_density"),
c("CO2_emissions", "Electricity", "Health_expenditure", "Out_of_pocket", "Private_health_expenditure", "Govt_health_expenditure",
  "Freshwater_resources", "HIV_prevalence", "Unemployment", "Govt_effectiveness", "Income_share", "Tobacco_use",
  "Alcohol_consumption","Political_stability", "Population_density"))
R_p

R_r <- cor(SIGMA_r, use = "pairwise.complete.obs")
R_r
```

### Eigenvalues for response/predictor

```{r}
predictor <- eigen(R_p)$values
predictor
```

### Intrinsic Dimensionality: Predictors

```{r}
sum(eigen(R_p)$values>1.0)

sum(eigen(R_p)$values>0.7)

plot(eigen(R_p)$values, type = "b")
```

We applied 3 types of criterion to the response variable.

For Kaiser's criterion, 5 values are retained since only 5 values are above 1.

For Jolliffe's criterion, 8 values are retained because they are above the threshold of 0.7.

We think that Kaiser's criterion and Jolliffe's criterion represents our dimensionality the best. The average of these is 6.5, which we will round to an intrinsic dimensionality of 7.

### Intrinsic Dimensionality: Response

Since we only have a since response dimension are intrinsic dimensionality will be 1 no matter what criterion is applied.

### Rotated Loading Matrix: Orthogonal

```{r}
A <- pca(r = R_p, nfactors = 7, rotate = "varimax")$loadings[]
A
corrplot(A)
```

### Rotated Loading Matrix: Oblique

```{r}
A_1 <- pca(r = R_p, nfactors = 7, rotate = "oblimin")$loadings[]
A_1
F_1 <- pca(r = R_p, nfactors = 7, rotate = "oblimin")$Phi
F_1
corrplot(A_1)
```

There are correlations among factors above 0.3 , so oblique rotation is indeed necessary.

### Communality

```{r}
pca(r = R_p, nfactors = 7, rotate = "oblimin")$communality
```

For a seven factor solution, all the dimensions retain above 70% of the variance of the original dimensions. This is a good because we avoid excessive data loss.

### Variance

```{r}
pca(r = R_p, nfactors = 7, rotate = "oblimin")
```

For a seven factor solution,

-   TC7 accounts for 19% of the entire variance

-   TC1 accounts for 18% of the entire variance

-   TC6 accounts for 16% of the entire variance

-   TC2 accounts for 13% of the entire variance

-   TC3 accounts for 13% of the entire variance

-   TC5 accounts for 11% of the entire variance

-   TC4 accounts for 10% of the entire variance
